Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"} \footnote{Funded by the Swedish Research Council under grants 349-2008-6592 and 421-2012-892.}

### 2018-10-08 popstud_open.Rmd
fix abstract and acknowledgement

\section*{Acknowledgement}

We thank the Demographic Data Base, Umeå University, for giving us access to data and
Jonas Helgertz at the Department of Economic History, Lund University, for
providing us with scripts for creating Hisclass from Swedish census
data.

@@

<!--
```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4,cache=FALSE}
cls <- rainbow(5, s = 1.0)
cls <- cls[c(2, 3, 1, 5, 4)]
cls[1] <- "black"
oldcols <- c("red", "darkgreen", "blue", "green", "black")
xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4,cache=FALSE}
xx2 <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy2 <- round(prop.table(xx2, 1) * 100, 1)
barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
```
-->

@@

```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4}
xx <- with(vb, tapply(exposure, list(period = period, civst = civst), sum))
##xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
##xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[3:1, ], col = c("red", "blue", "green"), las = 2, ylab = "Per cent")
```
-->

@@

```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4}
orr <- aggregate(x = vb$exposure, by = vb[, c("period", "civst", "hisclass")], FUN = sum)

@@

```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4}
orr <- aggregate(x = vb$exposure, by = vb[, c("period", "urban", "hisclass")], FUN = sum)

@@

```{r exweight,echo=TRUE,echo = FALSE}
vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
fit1 <- glm(I(hisclass == "elite") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit2 <- glm(I(hisclass == "middle") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit3 <- glm(I(hisclass == "worker") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit4 <- glm(I(hisclass == "farmer") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit5 <- glm(I(hisclass == "none") ~ urban * civst, data = vb2_2005_men, family = binomial)
wght <- numeric(NROW(vb2_2005_men))
##
gg <- vb2_2005_men$hisclass == "elite"
wght[gg] <- 1 / predict(fit1, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "middle"
wght[gg] <- 1 / predict(fit2, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "worker"
wght[gg] <- 1 / predict(fit3, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "farmer"
wght[gg] <- 1 / predict(fit4, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "none"
wght[gg] <- 1 / predict(fit5, type = "response")[gg]
##
vb2_2005_men$wght <- wght
```

@@

```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4}
par(cex.axis = 0.8)
fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = vb2_2005_men)
med <- summary(survfit(fit))$table[, "median"]
fitw <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = vb2_2005_men)
medw <- summary(survfit(fitw))$table[, "median"]
oldpar <- par(mfrow = c(1, 2))
plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100),
     xlab = "Age", ylab = "Survival", main = "Unweighted", axes = FALSE)
tics <- c(65, round(med["none"], 1), round(med["elite"], 1), 100)
axis(1, at = tics, las = 1, cex.axis = 0.8)
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(h = 0)
abline(h = 0.5, lty = 3)

@@

```{r estmed}
knitr::kable(round(medi, 1), caption = "Estimated medians, men 65-99, 2005-2013.")
```

@@

```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4}
source("R/getWeights.R")
da <- vb1[vb1$sex == "male" & vb1$period == "2005-2013", ]
da$wght <- getWeights(da)
##
fit0 <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = da)
med0 <- summary(survfit(fit))$table[, "median"]
fitw0 <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = da)
medw0 <- summary(survfit(fitw))$table[, "median"]
oldpar <- par(mfrow = c(1, 2))
plot(survfit(fitw0), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65),
     xlab = "Age", ylab = "Survival", main = "40-65", axes = FALSE)
tics <- c(40, 45, 50, 55, 60, 65)
axis(1, at = tics)
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(h = 0)
abline(h = 0.5, lty = 3)
legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.7)
##lines(c(med["none"], med["none"]), c(0, 0.5), col = "red", lty = 3)
##lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100),
     xlab = "Age", ylab = "Survival", main = "65-100", axes = FALSE)
tics <- c(65, 70, 75, 80, 85, 90, 95, 100)
axis(1, at = tics)
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(h = 0)
abline(h = 0.5, lty = 3)
legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.7)
```

@@

<!--
```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass. Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- resw[3, ]
plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2015),
     ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)

>>>>

\section*{Acknowledgement}
\thispagestyle{empty}
We thank the Demographic Data Base, Umeå University, for giving us access to data and
Jonas Helgertz at the Department of Economic History, Lund University, for
providing us with scripts for creating Hisclass from Swedish census
data.

@@

<!--
```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4,cache=FALSE}
cls <- rainbow(5, s = 1.0)
cls <- cls[c(2, 3, 1, 5, 4)]
cls[1] <- "black"
oldcols <- c("red", "darkgreen", "blue", "green", "black")
xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
##barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4,cache=FALSE}
xx2 <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy2 <- round(prop.table(xx2, 1) * 100, 1)
##barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
```
-->

@@

```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4,eval = FALSE}
xx <- with(vb, tapply(exposure, list(period = period, civst = civst), sum))
##xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
##xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[3:1, ], col = c("red", "blue", "green"), las = 2, ylab = "Per cent")
```
-->

@@

```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4,eval=FALSE}
orr <- aggregate(x = vb$exposure, by = vb[, c("period", "civst", "hisclass")], FUN = sum)

@@

```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4,eval=FALSE}
orr <- aggregate(x = vb$exposure, by = vb[, c("period", "urban", "hisclass")], FUN = sum)

@@

```{r exweight,echo=TRUE,echo = FALSE,eval=FALSE}
vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
fit1 <- glm(I(hisclass == "elite") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit2 <- glm(I(hisclass == "middle") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit3 <- glm(I(hisclass == "worker") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit4 <- glm(I(hisclass == "farmer") ~ urban * civst, data = vb2_2005_men, family = binomial)
fit5 <- glm(I(hisclass == "none") ~ urban * civst, data = vb2_2005_men, family = binomial)
wght <- numeric(NROW(vb2_2005_men))
##
gg <- vb2_2005_men$hisclass == "elite"
wght[gg] <- 1 / predict(fit1, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "middle"
wght[gg] <- 1 / predict(fit2, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "worker"
wght[gg] <- 1 / predict(fit3, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "farmer"
wght[gg] <- 1 / predict(fit4, type = "response")[gg]
##
gg <- vb2_2005_men$hisclass == "none"
wght[gg] <- 1 / predict(fit5, type = "response")[gg]
##
vb2_2005_men$wght <- wght
```

@@

```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4,eval=FALSE}
par(cex.axis = 0.8)
fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = vb2_2005_men)
med <- summary(survfit(fit))$table[, "median"]
fitw <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = vb2_2005_men)
medw <- summary(survfit(fitw))$table[, "median"]
oldpar <- par(mfrow = c(1, 2))
plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100),
     xlab = "Age", ylab = "Survival", main = "Unweighted", axes = FALSE)
tics <- c(65, round(med["none"], 1), round(med["elite"], 1), 100)
axis(1, at = tics, las = 1, cex.axis = 0.8)
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(h = 0)
abline(h = 0.5, lty = 3)

@@

```{r estmed,eval=FALSE}
knitr::kable(round(medi, 1), caption = "Estimated medians, men 65-99, 2005-2013.")
```

@@

```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4,eval=FALSE}
source("R/getWeights.R")
da <- vb1[vb1$sex == "male" & vb1$period == "2005-2013", ]
da$wght <- getWeights(da)
##
fit0 <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = da)
med0 <- summary(survfit(fit))$table[, "median"]
fitw0 <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = da)
medw0 <- summary(survfit(fitw))$table[, "median"]
oldpar <- par(mfrow = c(1, 2))
plot(survfit(fitw0), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65),
     xlab = "Age", ylab = "Survival", main = "40-65", axes = FALSE)
tics <- c(40, 45, 50, 55, 60, 65)
axis(1, at = tics)
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(h = 0)
abline(h = 0.5, lty = 3)
legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.7)
##lines(c(med["none"], med["none"]), c(0, 0.5), col = "red", lty = 3)
##lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100),
     xlab = "Age", ylab = "Survival", main = "65-100", axes = FALSE)
tics <- c(65, 70, 75, 80, 85, 90, 95, 100)
axis(1, at = tics)
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(h = 0)
abline(h = 0.5, lty = 3)
legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.7)
```

@@

<!--
```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass. Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm",eval=FALSE}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- resw[3, ]
plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2015),
     ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
